### 2020-02-21 Writing/disparity_review.tex
added figures to tex + compile OK

%Packages
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{pdflscape}

@@

Analyses of morphological disparity have been used to characterise and investigate the evolution of variation in the anatomy, function, and ecology of organisms since the 1980s.
While a diversity of methods have been employed, it is unclear whether they provide
equivalent insights.
Here we review the most commonly used approaches for characterising and analysing morphological disparity, all of which have associated limitations that, if ignored, can lead to misinterpretation.
We provide best practice guidelines for disparity analyses, while noting that there can be no ``one-size-fits-all'' approach.
The available tools should always be used in the context of a specific biological question that will determine data and method selection at every stage of the analysis.

@@

Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids and molluscs), species-rich but phenotypically conservative (e.g. bacteria and nematodes), species-poor but phenotypically diverse (e.g. Afrotheria), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon. Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

@@

\textit{Figure disparities {[}figure\_disparities.pdf{]}: The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
These categories are not independent and many disparity studies will cover more than one.}

%@@@ Add disparity and evo-devo

@@

Disparity analyses are based on traits, but traits can be characterised in a number of ways:
% MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g. Raup's shell coiling models). I'm not convinced that we must, but it may come across as an notable omission to reviewers
% SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
	1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Close2015-qi};
	2) continuous measurements of features (e.g., lengths) \citep[][e.g]{Anderson2001-qb}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, and Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.

@@

%\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.

\textit{Figure data {[}figure\_data.pdf{]}: Major routes to obtain morphological data for disparity analyses. Data can be collected as discrete trait observations (e.g. presence or absence data) or as continuous data.
Continuous data can be collected by various methods including linear measurements and landmark coordinates or contours (curves).
These measurements can then be mathematically transformed (logarithm transforms, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.}

\section{Disparity methods}

Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the variety of ordination (dimension reduction) techniques available and their limitations; (c) the metrics used to summarise the relative disparity of groups; (d) the methods used for hypothesis testing within the disparity analysis framework; and (e) ancestral state estimations in disparity-through-time analyses.
% @@@ Check the numbering here
We consider these aspects in turn below.

@@

%\includegraphics[width=6.5in,height=5.26389in]{media/image3.png}

\textit{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
% SG: The morphospace is the original trait space within which lie the specimens. Then one performs an ordination to visualize the space. So I would distinguish the morphospace from the ordination space. See also Mitteroecker and Huttegger (2009)
A trait matrix can be transformed into a distance matrix (e.g. in Close et al., 2015) or an ordinated matrix (e.g. in Brusatte et al.} \emph{2008).
In each case, all three matrices represent the morphospace for the observations at hand.
Visualisation: different ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling} \emph{forcing all the variance to be contained in two dimensions)
% SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
; or ordination axis plots (directly from the ordinated matrix).}
% ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
% SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
% @@@ Change references

@@

\section{4. Disparity analyses for the future}

Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse it, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate the bespoke
%VW: I think there could be just a line or two that takes makes "bespoke design" less daunting. There are loads of well-documented R packages available in morphological and ecological studies that can be relatively straightforwardly adapted.
assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets
\citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}
. % @@@ Add Mammola
Innovations in morphological disparity analyses likely await discovery in their respective literatures.
% AG: as above, i think you need to add in evodevo work here, it's really fundamental to disparity

@@

\subsection{Acknowledgments}

@@

>>>>

%Packages
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{pdflscape}

@@

\noindent Analyses of morphological disparity have been used to characterise and investigate the evolution of variation in the anatomy, function, and ecology of organisms since the 1980s.
While a diversity of methods have been employed, it is unclear whether they provide
equivalent insights.
Here we review the most commonly used approaches for characterising and analysing morphological disparity, all of which have associated limitations that, if ignored, can lead to misinterpretation.
We provide best practice guidelines for disparity analyses, while noting that there can be no ``one-size-fits-all'' approach.
The available tools should always be used in the context of a specific biological question that will determine data and method selection at every stage of the analysis.

@@

\noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids and molluscs), species-rich but phenotypically conservative (e.g. bacteria and nematodes), species-poor but phenotypically diverse (e.g. Afrotheria), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon. Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
\caption{
    The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
    These categories are not independent and many disparity studies will cover more than one.
    %@@@ Add disparity and evo-devo
}
\label{Fig:disparity}
\end{figure}

@@

\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
% MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g. Raup's shell coiling models). I'm not convinced that we must, but it may come across as an notable omission to reviewers
% SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
	1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Close2015-qi};
	2) continuous measurements of features (e.g., lengths) \citep[][e.g]{Anderson2001-qb}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, and Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_data.pdf}
\caption{
    Major routes to obtain morphological data for disparity analyses. Data can be collected as discrete trait observations (e.g. presence or absence data) or as continuous data.
    Continuous data can be collected by various methods including linear measurements and landmark coordinates or contours (curves).
    These measurements can then be mathematically transformed (logarithm transforms, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
    Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.
}
\label{Fig:data}
\end{figure}

\section{Disparity analysis methods}

\noindent Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the variety of ordination (dimension reduction) techniques available and their limitations; (c) the metrics used to summarise the relative disparity of groups; (d) the methods used for hypothesis testing within the disparity analysis framework; and (e) ancestral state estimations in disparity-through-time analyses.
% @@@ Check the numbering here
We consider these aspects in turn below.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
\caption{
    Morphospaces: different mathematical representations of a morphospace.
    % SG: The morphospace is the original trait space within which lie the specimens. Then one performs an ordination to visualize the space. So I would distinguish the morphospace from the ordination space. See also Mitteroecker and Huttegger (2009)
    A trait matrix can be transformed into a distance matrix \citep[][e.g in]{Close2015-qi} or an ordinated matrix \citep[][e.g. in]{Brusatte2008-vx}
    In each case, all three matrices represent the morphospace for the observations at hand.
    Visualisation: different ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling forcing all the variance to be contained in two dimensions)
    % SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
    ; or ordination axis plots (directly from the ordinated matrix).
    % ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
    % SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
    % @@@ Change references
}
\label{Fig:morphospace}
\end{figure}

@@

\section{Disparity analyses for the future}

\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse it, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate the bespoke
%VW: I think there could be just a line or two that takes makes "bespoke design" less daunting. There are loads of well-documented R packages available in morphological and ecological studies that can be relatively straightforwardly adapted.
assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets
\citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}
. % @@@ Add Mammola
Innovations in morphological disparity analyses likely await discovery in their respective literatures.
% AG: as above, i think you need to add in evodevo work here, it's really fundamental to disparity

@@

\subsection{Acknowledgements}

@@

\bibliographystyle{sysbio}
\bibliography{References}
